---
title: "Unit 9: Multiple predictors"
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
format:
  clean-revealjs:
    self-contained: true
    chalkboard: false
    standalone: true
    slide-number: true
    show-slide-number: print
code-annotations: hover
bibliography: QMLS_Bibliography.bib
csl: evolution.csl
knitr:
  opts_chunk: 
    echo: true
---

```{r}
#| label: setup
#| echo: false

library(tidyverse)
library(performance)
library(multcomp)
library(readxl)
library(GGally)
library(wesanderson)
library(ggrepel)
library(ggforce)
library(plotly)

source("https://raw.githubusercontent.com/QM-Life-Sci/QMLS_1/refs/heads/main/Lectures/QMLS_functions.R")

theme_set(theme_classic(base_size = 20))
```


# Housekeeping {background-color="#40666e"}

- PC 2 check-in
    - Due Thursday. Let me know if you need more time.
- Data sources for micropublication
    - [Dryad](https://datadryad.org/)
    - [Figshare](https://figshare.com/browse)
    - [Zenodo](https://zenodo.org/)
    - [List of repositories](https://www.nature.com/sdata/policies/repositories)


# Multicollinearity {background-color="#40666e"}

## Multicollinearity

High correlation between predictors leaves little residual variation to be used for explaining the outcome variable.

```{r}
ggscatmat(Milk, columns = 2:4)
```


## How much correlation is too much?

- No universal guidelines, much discussion:
    - Farrar and Glauber [-@Farrar1967-mn]
    - Graham [-@Graham2003-yz]
    - Tu et al. [-@Tu2004-va]
    - Vatcheva et al. [-@Vatcheva2016-tr]
- Correlation (*r*) > 0.8 should get a closer look (some say 0.5)
- Variance inflation factor (VIF): inflation of the parameter estimate variances due to shared variance


## Diagnostics

```{r}
#| echo: true
#| eval: false

performance::check_model(fm)
```

```{r}
print(PP)
```


## Masking

Multiple predictors are useful for predicting outcomes when bivariate relationships with the response variable is not strong.

But:

- Associative relationships can be obscured when two predictors are somewhat correlated with one another.


## Milk is a large energetic investment

- Is there a significant association between energy content of milk while controlling for neocortex size and body size?
- Do primates with larger brains produce significantly more nutritious milk so their offspring can grow quickly (because they must grow quickly) ?


## Visualizing

```{r}
Milk <- read_excel("../data/Milk.xlsx", na = "NA") |> 
  select(Species, Milk_Energy, Mass, Neocortex) |>
  mutate(log_Mass = log(Mass)) |> 
  drop_na()

p1 <- ggplot(Milk, aes(log_Mass, Milk_Energy)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  labs(x = "log Mass (kg)", y = "Milk Energy (kcal/g)") +
  geom_text_repel(aes(label = Species), size = 3, force = 15)
p2 <- ggplot(Milk, aes(Neocortex, Milk_Energy)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  labs(x = "Percentage Neocortex", y = "Milk Energy (kcal/g)") +
  geom_text_repel(aes(label = Species), size = 3, force = 15)
plot_grid(p2, p1, ncol = 2)
```


## Visualizing

```{r}
ggscatmat(Milk, columns = c(2, 4, 5))
```


## Bivariate model of Neocortex

```{r}
#| echo: true

fm_Neo <- lm(Milk_Energy ~ Neocortex, data = Milk)
summary(fm_Neo)
```


## Bivariate model of log Mass

```{r}
#| echo: true

fm_Mass <- lm(Milk_Energy ~ log_Mass, data = Milk)
summary(fm_Mass)
```


## Multiple predictors model

```{r}
#| echo: true

fm_Multi <- lm(Milk_Energy ~ Neocortex + log_Mass, data = Milk)
summary(fm_Multi)
```


## Interpretation

- Both coefficients go up
    - Neocortex: $0.005 \rightarrow 0.03$ (*P* = 0.004)
    - log Mass: $-0.03 \rightarrow -0.1$ (*P* = 0.002)

Regression *simultaneously* asks (and answers):

1. Do species that have high neocortex percentage *for their mass* have higher energy milk?
1. Do species with high body mass *for their neocortex percentage* have higher energy milk?


## Neocortex vs. log Mass

```{r}
ggplot(Milk, aes(log_Mass, Neocortex)) +
  geom_point() +
  labs(x = "log Mass (kg)", y = "Neocortex Percentage") +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  geom_text_repel(aes(label = Species), size = 3, force = 15)
```


## Milk Energy vs. Residual Neocortex

```{r}
Milk$resid_Neo <- residuals(lm(Neocortex ~ log_Mass, data = Milk))
ggplot(Milk, aes(resid_Neo, Milk_Energy)) +
  geom_vline(xintercept = 0, color = "blue") +
  geom_point() +
  labs(x = "Residual Neocortex", y = "Milk Energy") +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE,
              color = "darkgreen") +
  geom_text_repel(aes(label = Species), size = 3, force = 15)
```


# Questions {background-color="#f8c471"}

