---
title: 'Problem Set 02'
author: "Your Name Here"
date: 'Last updated: `r Sys.Date()`'
format:
  html:
    toc: true
    toc_depth: 2
    toc_float: true
    embed-resources: true
---

```{r}
#| label: setup
#| message: false
#| warning: false

```

> "...the descriptive and analytical power of an adequate programming language amply repays the considerable effort required for its mastery." Ken Iverson^[Iverson, K. E. 1962. A Programming Language. John Wiley & Sons Inc.]


## How to use the problem sets

Each problem set will contain narrative text interspersed with R code. Some of this code will already be completed for you, while some you will need to fill in. You should read through all of the text (just as you are doing right now). The exercises (**Activities**) are interspersed throughout the text.

Right now, go up to the header and change the line "author:" to have your name.


### Important

As you work through the exercises, really think about what you are doing. In the upcoming problem sets, you will be learning how to think through analyses step by step, from loading data, to visualizing it, to actually carrying out statistical tests, and then interpreting the results.

Read the code like you would a textbook (meaning studying and digesting as you go). It's easy to copy and paste or just run code chunks without really thinking about what the code is doing (we've done it too), but that's not a good way to learn.


## Insert an R code chunk

An Rmd file can include R code that is run at compile time.


### Activity

**Activity** means that you need to do something (i.e., be active). Before you submit your problem set, search (*Ctrl-f* / *Cmd-f*) for "Activity" and make sure you have answered all the questions.

Place the cursor on the line below this text and press *Ctrl-Alt-i* (Windows) / *Cmd-Option-i* (Mac) to insert an R code chunk.



Enter some R code into the chunk on the blank line: `sqrt(2)`. Then compile the HTML. Your file will show the R code that is run and R's output ($\sqrt{2} = 1.41$).

You can also run code interactively from a Markdown file to the R console. To run a single line, press *Ctrl-Enter* / *Cmd-Return*. To run the current chunk, use *Ctrl-Alt-c* / *Cmd-Shift-Return* or click on the small green triangle in the upper-right corner of the chunk. This is a good way to test code that you are working on, rather then waiting to compile the file to HTML (or whatever format you are using).

You can also enter the code to set up code chunks manually, but we find it easier to use the insert code chunk shortcut.


## Naming R objects

There are not many restrictions on what you can name R objects. Unlike other languages, where some names are reserved and off-limits, in R, pretty much anything goes. Object names can't start with a number (i.e., no `1a <- 1`; that will give you an "unexpected symbol" error), but otherwise you are free to do what you want, even things you probably should not do. One of the main things to avoid is naming your objects the same as an R function.

Some names to avoid: `c`, `mean`, `df`, `matrix`, `t`, `T`, `F`. The last two are acceptable abbreviations for `TRUE` and `FALSE`. To avoid ambiguity, I recommend writing out `TRUE` and `FALSE` explicitly, rather than using the abbreviations.

If you want to take the mean of a vector `x`, `mean_x`, `x_bar`, or `x_mean` is preferred.^[I prefer the latter two because of RStudio's auto-completion feature.] There are two benefits of using one of these variable names over using `mean`.

1. You don't confuse your `mean` object with the `mean()` function. R will *usually* figure out which one you want, but we always encourage users to be explicit rather than relying on defaults.
2. What if you later want to take the mean of a different vector. Which one does `mean` refer to?

You could do this, for example:

```{r}
# Assign the standard deviation of the vector 1 to 6 to an object called sd.
sd <- sd(1:6)

# Print standard deviation of 1-6
sd

# Print standard deviation of 4-10
sd(4:10)
```


### Activity

Execute the chunk above and look at the R console output. Explain what we have done here and what R must be doing without telling you. Write your answer after the ">" below. (">" is the Markdown syntax for a block quote, which will make finding your answers easier.)

>


Now add comments to the code chunk above briefly annotating your answer above.


## Vectors

Vectors are one of the fundamental data structures in R. They consist of data of all the same type (numeric, character, etc.) in a 1 X n structure. You can manually create vectors using the combine function `c()`. Some functions like `seq()`, `rep()`, and the random number generators (`rnorm()`, `runif()`, etc.) produce vectors by default.


### Activity

Assign vectors with the following characteristics (for simplicity, you can call them `v1` through `v9`):

1. 1, 6, 10, 14.75
2. TRUE, TRUE, FALSE
3. a, aa, aaa (as characters)
4. The sequence 5 to 100 by 1
5. The sequence 5 to 100 by 5
6. The sequence starting with 5 and ending at 100 with a length of 60 (i.e., 60 evenly spaced values from 5 to 100)
7. 17 repeated 10 times
8. The sequence 1, 2, 3 where each is repeated 10 times in a row (i.e., 1, 1, 1, 1, ..., 2, 2, 2, 2, ..., 3, 3, 3, 3, ...)
9. The sequence 1, 2, 3 repeated 10 times (i.e., 1, 2, 3, 1, 2, 3, 1, 2, 3, ...)

Look at the help for `seq()` and `rep()` to help you with 5 through 9 (and maybe 4).

Choose names for each of these vectors. Add your code to the block below.

```{r}

```


## Working with relational operators

Binary operations are very important in R for selecting, subsetting, and choosing variables. The relational operators in R are:

- `==` Equals
- `!=` Does not equal
- `>` Greater than
- `<` Less than
- `>=` Greater than or equal to
- `<=` Less than or equal to
- `%in%` Is the comparator in the set?

When these operators are applied to vectors, the result is a vector of logicals (`TRUE`s and `FALSE`s).


### Activity

Use your vectors from above in the same order to test the following relational operators.

The numbers (1-9) in the list below correspond to those in the list above.

1. Which values are greater than 5?
2. Which values equal FALSE?
3. Does this vector contain the string "a"?
4. Which values are less than or equal to 10?
5. Which values are greater than or equal to 10?
6. Count the number of values less than 50 (hint, use `sum()`)
7. Count the number of values equal to 17
8. Which values equal 1?
9. Which values do not equal 1?

```{r}

```


## Perils of relational operators

Computers only infrequently store numbers internally as integers (although [computer algebra systems](https://en.wikipedia.org/wiki/Computer_algebra_system) do), particularly after any kind of numeric operation.

In contrast, numbers^[If you have experience with numeric types, we are talking about floating point numbers.] are often rounded to some level of accuracy (R uses about 53 decimal places). For example:

```{r}
a <- sqrt(2)
a
a * a
a * a == 2
all.equal(a * a, 2)
```

Line by line, explain what the statements above are doing and the R output of each. Look at the help for `all.equal()` if you need to. Enter your explanation after the `>` below.

>


## `data.frame()`s and `tibble()`s

`data.frame`s (and their cousins `tibble`s) are one of the most important objects that you will work with in R. They are the closest thing to an Excel spreadsheet in R (with the added restriction that a column must be all of one data type).


### Activity

Create a `data.frame` or `tibble` with the following columns:

- A character vector `Tx_Group` with the values "control" and "high", each repeated 10 times.
- A numeric vector `Replicate` with values 1 and 2, repeated 5 times each for both values in `Tx_Group`.
- A numeric vector `Productivity`, where the first 10 values are normally distributed with a mean of 5 and standard deviation of 2 and the second 10 values are normally distributed with a mean of 8 and standard deviation of 2 (look at the help for `rnorm()` for help). `c()` will be useful here.

If you use a `tibble()`, you will need to load the `tidyverse` package: `library(tidyverse)`. Typically, packages are loaded at the beginning of a script (e.g., the first `setup` chunk).

```{r}
library(tidyverse)

set.seed(15)

dd <- tibble(Tx_Group = rep(c("control", "high"), each = 10),
       Replicate = c(rep(c(1, 2), each = 5),
                     rep(c(1, 2), each = 5)),
       Productivity = c(rnorm(n = 10, mean = 5, sd = 2),
                        rnorm(n = 10, mean = 8, sd = 2))
       )

ggplot(data = dd, aes(x = Tx_Group, y = Productivity, color = factor(Replicate))) +
  geom_point()


```

Use the `str()` function to get information about the data.frame. This will allow you to verify that `Tx_Group` has the type character. Note that even though `Replicate` only contains the integers 1 and 2, R treats it as a numeric.

```{r}

```


## Indexing

Taking subsets of objects in R is very common. This can include slicing or filtering rows, extracting one or more columns, and referencing columns in other functions.

You can use standard bracket notation `[ ]` to subset vectors, matrices, and data.frames. The latter two require a comma to denote rows and columns: `[rows, columns]`. You can also take a single column of a data.frame with the `$` operator.


### Activity

Use your data.frame (or tibble) from the question above. Extract the following subsets:

1. The column `Productivity` using bracket notation
2. The column `Productivity` using `$` notation
3. The second column (assume you don't know its name)
4. Rows 1-10 of the entire data.frame
5. Rows 1-10 of only the Productivity column

```{r}

```

We will do more complex filtering next week (e.g., only rows from replicate 1 where productivity is less than 5).


## Basic calculations in R

R can do basic (and not so basic) calculations. First set up a vector of numbers to work with.

```{r r_calculator}
# Set the random number generator seed
set.seed(5)

# Generate 10 random numbers between 0 and 1
x <- runif(10)
x
```


### Activity

Try out some R functions: `mean(x)`, `sd(x)`, `median(x)`, `t(x)`. Type them into the code chunk below and compile.

These functions take a vector or matrix and return a single value or a new matrix. Contrast this behavior with `x^2`. Enter that and see what you get.

Try functions operating on the matrix you created above.

```{r}

```


## Cleaning up code

Spaces and returns are your friend. Imaginetryingtoreadasentencethathasnospacing. It's pretty difficult. Writing clean, organized code helps your future self and those you work with to understand the steps of your analysis.


### Activity

Take the following code chunk and make the code more readable by adding spaces and returns as well as explicitly naming some of the arguments (check the help files for the assumed ordering of argument).

```{r}
# FIXME
M1<-data.frame(x=11:20,y=c(20:11),l=letters[1:10])
M2 <- tibble(ID=seq(1,100,1),x=rnorm(100,10,5),y=8.4*x+rnorm(100,0,1))
```

What is `letters`?

>

Explain how `y` is constructed in `M2`.

>
