---
title: 'Problem Set 07'
author: "Your Name Here"
date: 'Last updated: `r Sys.Date()`'
format:
  html:
    toc: true
    toc_depth: 2
    toc_float: true
    embed-resources: true
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(performance)
library(see)
library(broom)
library(readxl)
library(cowplot)

theme_set(theme_classic())
```

## Setup

This problem set will use some additional packages that you might not have installed:

- "performance"
- "see"
- "qqplotr"
- "broom"
- "readxl"
- "cowplot"

If you do not have these packages installed already, go ahead and do that before you start. You can use the "Packages" pane in RStudio.


## Lion noses

In PS 6, we continued analyzing the data set where lion age was related to the proportion of black coloration in the nose. The goal was to be able to predict a lion's age from only a photograph of its nose.

### Activity

Copy your code from PS 6 to import the `LionAges.csv` and make a scatterplot of age regressed on proportion of black.

Add an OLS regression line to your plot using `geom_smooth()` following the examples in the lecture slides and PC 1.

```{r}
#| fig-alt: "A scatterplot of lion age versus the proportion black on the nose. A best fit regression line is overlaid."

# FIXME

LA <- read_csv("../data/LionAges.csv", show_col_types = FALSE)

ggplot(LA, aes(y = age, x = proportion.black)) +
  geom_point(size = 4, color = "navy") +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(breaks = 1:14) +
  labs(x = "Proportion Black", y = "Age (y)")
```

Copy over the linear model you fit where age is modeled by proportion of black.

```{r}
# FIXME

fm <- lm(age ~ proportion.black, data = LA)
```

Pass the `lm()` object to the `check_model()` function. This will print a 5-panel figure of diagnostics. You can use the "Zoom" button in the Plots panel to make the image larger.

```{r}
#| fig-height: 10
#| fig-alt: "A set of 5 plots displaying the model diagnostics for posterior predictive check, linearity, homogeneity of variance, influential observations, and normality of residuals."

# FIXME

check_model(fm)
```

We'll go through the diagnostics one-by-one.

What does the "Posterior Predictive Check" represent? Do you see any problems with it?

> The green line is the density plot of the observed y values. The blue lines are simulated values based on the model fit. This looks fine here.


What does the "Linearity" panel check? Do you see any problems with it?

> This is a check that there really is a linear relationship between the x and y variables. The line is pretty wavy, but there are also not a lot of data points. It seems flat enough.


What does the "Homogeneity of Variance" panel check? Do you see any problems with it?

> This is a check that the variation in the residuals is (relatively) constant across all the values of x. It's mostly flat here, but it does deviate away fro, horizontal at large fitted values of age.


What does the "Influential Observations" panel check? Do you see any problems with it?

> This panel shows a diagnostic for points that have an undue influence on the linear model (e.g., high "leverage"). This plot show that point #30 is influential.


Finally, run the three additional checks:

- `check_heteroskedasticity()`
- `check_normality()`
- `check_outliers()`

```{r}
# FIXME

check_heteroskedasticity(fm)
check_normality(fm)
check_outliers(fm)
```

What do these checks report? Do the results agree with the diagnostic plots?

> `check_heteroskedasticity()` reports significant non-homogeneity, which appears to agree with plot #3, where the variance increases with the fitted values. `check_outliers()` reports that observation 30 is an outlier.


In class we made a plot using `geom_label()`, with the observation numbers plotted instead of points. Adapt that plot here to identify the influential observation. Include the code you wrote above to also draw the OLS regression line on the plot.

```{r}
#| fig-alt: "A scatterplot of lion age versus the proportion black on the nose. A best fit regression line is overlaid. Points are displayed with their index number."

# FIXME

LA |>
  mutate(ID = factor(1:n())) |>
  ggplot(aes(proportion.black, age)) +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  geom_label(aes(label = ID)) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(breaks = 1:14) +
  labs(x = "Proportion Black", y = "Age (y)")
```

What does it mean to say that point #30 has a lot of *influence* on the regression as compared to, for example point #17?

> Point 30 is both away from the regression line and at an extreme value of proportion black. Thus it is able to exert a significant "pull" on the regression line. Point 17, falls almost exactly on the regression line. So it is not able to pull the line much at all (the observed value is right where the model predicts it to be).

Filter out observation #30 by whatever method you want. Then refit the model and print the summary.

```{r}
# FIXME

# There are lots of options for dropping a single observation. Here we
# just use the fact that #30 has the largest value for age
LA_2 <- LA |> slice(-which.max(LA$age))
fm_2 <- lm(age ~ proportion.black, data = LA_2)
summary(fm_2)
summary(fm)
```

Compare the summaries for the original model with all the data points and the new model with the outlier removed. How much do the coefficients change? Does dropping one data point change your conclusions about the ability to predict age from proportion of black in the nose?

> The slope goes from 10.6 in the first model to 8.8 in the next model, so much lower, but still really high. The intercept changes as well, but that is not as interesting. Regardless of the change in slope, proportion.black is still a significant predictor of age. The P-value are both in the 10^-7 range.


## Waist Circumference and Serum Triglycerides

Cardiovascular disease is a leading cause of mortality in the United States. Dyslipidemia, in particular elevated serum triglycerides, is a well-established  risk factor for cardiovascular disease.

High serum triglycerides is associated with general metabolic dysregulation (insulin resistance, high blood pressure, Type 2 diabetes, and other physiological changes) and is one of the diagnostic criteria for "metabolic syndrome" (Grundy et al., 2004)^[Grundy, SM and others. (2004). Definition of metabolic syndrome: Report of the National Heart, Lung, and Blood Institute/American Heart Association conference on scientific issues related to definition. *Circulation* 109: 433-438.].

Visceral adipose tissue releases free fatty acids into circulation, leading to elevated serum triglycerides. Waist circumference is a easy-to-measure clinical measure of visceral adiposity.

In this question we want to explore the association between waist circumference and triglyceride levels and see whether triglyceride levels can be predicted from waist circumference. Because waist circumference is easy and free to measure, being able to predict serum triglycerides waist circumference would be beneficial.

The file `NHANES_Triglycerides.xlsx` is an Excel file that contains data with several body measurement and laboratory results from the [National Health and Nutrition Examination Survey ("NHANES")](https://wwwn.cdc.gov/nchs/nhanes/Default.aspx) in 2017-2018.

The columns in this data set are:

- `ID`: Participant's ID
- `Age`: Age in years
- `Sex`: Self-reported sex
- `BMI`: Body mass index in kg/m^2^
- `Waist`: Waist circumference in cm
- `Triglycerides`: Serum triglycerides in mg/dL

Use the function `read_excel()` from the package `readxl` to load the data and select only the self-reported females from the dataset.

```{r}
Tri <- read_excel("../data/NHANES_Triglycerides.xlsx") |>
  filter(Sex == "Female")
```

To explore the data, make two histograms, one for waist circumference and one for triglycerides.

```{r}
ggplot(Tri, aes(Waist)) +
  geom_histogram(bins = 30) +
  labs(x = "Waist Circumference (cm)")
```

```{r}
ggplot(Tri, aes(Triglycerides)) +
  geom_histogram(bins = 30) +
  labs(x = "Serum Triglycerides (mg/dL)")
```

What can you infer from the distributions of waist circumference and for triglycerides? Do these variable appear to be roughly normally distributed? do you have any concerns about using these data in a linear model?

> For waist circumference, the mean appears to be a bit less than 100 with a range from ~60 to ~160. There is some skew to the right with a few high values but it is not too severe. For triglycerides, the mean appears to be near 100 with a range from 0 to over 600. There is a stronger skew to the right with a few very high values that probably pull the mean out. There should be some concern using these data in a linear model because of this skew. 

Now explore the relationship between these variables by making a scatterplot of triglycerides versus waist circumference. Choose a color and size for your points and add informative axis labels.


```{r}
ggplot(Tri, aes(x = Waist, y = Triglycerides)) +
  geom_point() +
  labs(x = "Waist Circumference (cm)", y = "Serum Triglycerides (mg/dL)")
```

Perform an OLS regression predicting triglycerides from waist circumference.

```{r}
fm1 <- lm(Triglycerides ~ Waist, data = Tri)
```

Use `check_model` to check the diagnostics for your model.

```{r}
#| fig-height: 10

check_model(fm1)
```

Do you have any concerns about whether these data meet the assumptions for OLS regression? If any diagnostics are concerning, explain what the problem appears to be.

> Yes, I have some concerns. One concern is the posterior predictive check, where the distributions appear pretty different from the observed data and there is a lack of overlap as a result. In addition, the normality of residuals deviates pretty strongly from a horizontal line, likely because of the strong skew we noted previously. 


Transformations are often used to help meet the assumptions of our statistical tests. Let's try a log transformation of the triglycerides variable to see how this affects the model.

Create a new column that holds the (natural) log of triglycerides. Remake the scatterplot and perform the OLS regression this time with the log transformed variable.

```{r}
Tri <- Tri |> mutate(log_Tri = log(Triglycerides))

ggplot(Tri, aes(x = Waist, y = log_Tri)) +
  geom_point()

fm2 <- lm(log_Tri ~ Waist, data = Tri)
```

Now get the model summary and perform the diagnostics using check_model on this new regression.

```{r}
#| fig-height: 10

check_model(fm2)
```

```{r}
summary(fm2)
```

Compare the diagnostics for the log transformed data versus un-transformed. Which model would you recommend or not? Use specific diagnostic results to support your answer.

> The diagnostics are improved for the log transformed data. In particular the two checks we discussed previously have improved by a lot. The posterior predictive check now overlaps the observed data well and the normality of residuals looks as expected with a flat line. I would recommend proceeding with the log-transformed data for the model because of these improved diagnostics. 


Finally, using your plot above as a starter, make a publication-ready scatter plot of the log transformed triglycerides versus waist circumference.

- Add color and transparency as you see fit.
- Also add a line of best fit and choose a color for the line.
- Add informative axis labels.

```{r}
ggplot(Tri, aes(x = Waist, y = log_Tri)) +
  geom_point(color = "#7B1818", alpha = 0.5) +
  geom_smooth(
    formula = y ~ x,
    method = "lm",
    se = FALSE,
    color = "darkblue",
    linewidth = 2
  ) +
  labs(x = "Waist Circumference (cm)", y = "log Serum Triglycerides (mg/dL)")

```



## Stalk-eyed flies

In Problem Set 6 we continued analyzing the stalk-eyed fly data. We want to look at the diagnostics for that analysis now.

### Activity

Import the `Stalkies.csv` as in PS 6 and run the same linear model. Feel free to copy your code or the code from the key.

```{r}
# FIXME

stalk <- read_csv("../data/Stalkies.csv", show_col_types = FALSE) |>
  mutate(food_source = factor(food_source))

fm <- lm(eye_span ~ food_source, data = stalk)
summary(fm)
```


Pass the fitted model to `check_model()` as above to plot the regression diagnostics.

```{r}
#| fig-height: 10
#| fig-alt: "A set of 5 plots displaying the model diagnostics for posterior predictive check, linearity, homogeneity of variance, influential observations, and normality of residuals."

# FIXME

check_model(fm)

```

What do you notice is different about these diagnostic plots compared to the one above?

> There are only two x "values", so the plots look really wonky.


What can you conclude about the utility of diagnostic plots when the predictor is a categorical variable?

> It's not that useful.


Run the same 3 additional checks as above.

```{r}
# FIXME

check_heteroskedasticity(fm)
check_normality(fm)
check_outliers(fm)
```

What do these results show?

> These check so significant heteroskedasticity and non-normality of residuals.


Let's further investigate. In the chunk below, add a column to your stalk-eyed fly tibble with the residuals. There are a few ways to access the residuals:

- `fm$residuals`
- `resid(fm)`
- `augment()` from the `broom` package (see example in the lecture slides)

```{r}
# FIXME

stalk <- augment(fm, stalk)
```

Finally, plot a histogram of the *residuals* for the fitted model. Color by food source and facet the plot by food source.

```{r}
#| fig-alt: "A set of two histograms stacked vertically showing the distribution of residuals for the corn and cotton food source."

# FIXME

ggplot(stalk, aes(.resid, fill = food_source)) +
  geom_histogram() +
  facet_grid(food_source ~ .)
```

What do you notice about the residuals? Does this agree with the diagnostics above?

> The corn residuals are very compact and probably normally distributed. The cotton residuals are much more spread out, and it's hard to tell of they are normal or not. This seems to agree with the diagnostics: different variances and non-normal distributions of residuals.
