---
title: 'Problem Set 07'
author: "Your Name Here"
date: 'Last updated: `r Sys.Date()`'
format:
  html:
    toc: true
    toc_depth: 2
    toc_float: true
    embed-resources: true
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(performance)
library(see)
library(broom)
library(readxl)
library(cowplot)


theme_set(theme_classic())
```

## Setup

This problem set will use some additional packages that you might not have installed:

- "performance"
- "see"
- "qqplotr"
- "broom"
- "readxl"
- "cowplot"

If you do not have these packages installed already, go ahead and do that before you start. You can use the "Packages" pane in RStudio.


## Lion noses

In PS 6, we continued analyzing the data set where lion age was related to the proportion of black coloration in the nose. The goal was to be able to predict a lion's age from only a photograph of its nose.

### Activity

Copy your code from PS 6 to import the `LionAges.csv` and make a scatterplot of age regressed on proportion of black.

Add an OLS regression line to your plot using `geom_smooth()` following the examples in the lecture slides and PC 1.

```{r}

```

Copy over the linear model you fit where age is modeled by proportion of black.

```{r}

```

Pass the `lm()` object to the `check_model()` function. This will print a 5-panel figure of diagnostics. You can use the "Zoom" button in the Plots panel to make the image larger.

```{r}

```

We'll go through the diagnostics one-by-one.

What does the "Posterior Predictive Check" represent? Do you see any problems with it?

> 

What does the "Linearity" panel check? Do you see any problems with it?

> 

What does the "Homogeneity of Variance" panel check? Do you see any problems with it?

> 

What does the "Influential Observations" panel check? Do you see any problems with it?

> 


Finally, run the three additional checks:

- `check_heteroskedasticity()`
- `check_normality()`
- `check_outliers()`

```{r}

```

What do these checks report? Do the results agree with the diagnostic plots?

> 

In class we made a plot using `geom_label()`, with the observation numbers plotted instead of points. Adapt that plot here to identify the influential observation. Include the code you wrote above to also draw the OLS regression line on the plot.

```{r}

```

What does it mean to say that point #30 has a lot of *influence* on the regression as compared to, for example point #17?

> 


Filter out observation #30 by whatever method you want. Then refit the model and print the summary.

```{r}

```

Compare the summaries for the original model with all the data points and the new model with the outlier removed. How much do the coefficients change? Does dropping one data point change your conclusions about the ability to predict age from proportion of black in the nose?

> 

## Waist Circumference and Serum Triglycerides

Cardiovascular disease is a leading cause of mortality in the United States. Dyslipidemia, in particular elevated serum triglycerides, is a well-established  risk factor for cardiovascular disease.

High serum triglycerides is associated with general metabolic dysregulation (insulin resistance, high blood pressure, Type 2 diabetes, and other physiological changes) and is one of the diagnostic criteria for "metabolic syndrome" (Grundy et al., 2004)^[Grundy, SM and others. (2004). Definition of metabolic syndrome: Report of the National Heart, Lung, and Blood Institute/American Heart Association conference on scientific issues related to definition. *Circulation* 109: 433-438.].

Visceral adipose tissue releases free fatty acids into circulation, leading to elevated serum triglycerides. Waist circumference is a easy-to-measure clinical measure of visceral adiposity.

In this question we want to explore the association between waist circumference and triglyceride levels and see whether triglyceride levels can be predicted from waist circumference. Because waist circumference is easy and free to measure, being able to predict serum triglycerides waist circumference would be beneficial.

The file `NHANES_Triglycerides.xlsx` is an Excel file that contains data with several body measurement and laboratory results from the [National Health and Nutrition Examination Survey ("NHANES")](https://wwwn.cdc.gov/nchs/nhanes/Default.aspx) in 2017-2018.

The columns in this data set are:

- `ID`: Participant's ID
- `Age`: Age in years
- `Sex`: Self-reported sex
- `BMI`: Body mass index in kg/m^2^
- `Waist`: Waist circumference in cm
- `Triglycerides`: Serum triglycerides in mg/dL

Use the function `read_excel()` from the package `readxl` to load the data and select only the self-reported females from the dataset.

```{r}
NH <- read_excel("../data/NHANES_Triglycerides.xlsx") |> 
  filter(Sex == "Female")
```

To explore the data, make two histograms, one for waist circumference and one for triglycerides.

```{r}
ggplot(NH, aes(Waist)) +
  geom_histogram(bins = 30)
```

```{r}
ggplot(NH, aes(Triglycerides)) +
  geom_histogram(bins = 50)

NH |> arrange(desc(Triglycerides))

log(30)
exp(3.40)

log(3)
log(300)

ggplot(NH, aes(log(Triglycerides))) +
  geom_histogram(bins = 50)

```

What can you infer from the distributions of waist circumference and for triglycerides? Do these variable appear to be roughly normally distributed? do you have any concerns about using these data in a linear model?

>

Now explore the relationship between these variables by making a scatterplot of triglycerides versus waist circumference. Choose a color and size for your points and add informative axis labels.


```{r}
ggplot(NH, aes(x = Waist, y = Triglycerides)) +
  geom_point(color = "orchid") +
  geom_smooth(method = "lm", formula = y ~ x, se=FALSE)
```

Perform an OLS regression predicting triglycerides from waist circumference.

```{r}
fm <- lm(Triglycerides ~ Waist, data = NH)
summary(fm)
```

Use `check_model` to check the diagnostics for your model.

```{r}
check_model(fm)
```

Do you have any concerns about whether these data meet the assumptions for OLS regression? If any diagnostics are concerning, explain what the problem appears to be.

>


Transformations are often used to help meet the assumptions of our statistical tests. Let's try a log transformation of the triglycerides variable to see how this affects the model.

Create a new column that holds the (natural) log of triglycerides. Remake the scatterplot and perform the OLS regression this time with the log transformed variable.

```{r}

```

Now get the model summary and perform the diagnostics using check_model on this new regression.

```{r}

```

```{r}

```

Compare the diagnostics for the log transformed data versus un-transformed. Which model would you recommend or not? Use specific diagnostic results to support your answer.

>


Finally, using your plot above as a starter, make a publication-ready scatter plot of the log transformed triglycerides versus waist circumference.

- Add color and transparency as you see fit.
- Also add a line of best fit and choose a color for the line.
- Add informative axis labels.

```{r}


```



## Stalk-eyed flies

In Problem Set 6 we continued analyzing the stalk-eyed fly data. We want to look at the diagnostics for that analysis now.

### Activity

Import the `Stalkies.csv` as in PS 6 and run the same linear model. Feel free to copy your code or the code from the key.

```{r}

```


Pass the fitted model to `check_model()` as above to plot the regression diagnostics.

```{r}


```

What do you notice is different about these diagnostic plots compared to the one above?

> 


What can you conclude about the utility of diagnostic plots when the predictor is a categorical variable?

> 


Run the same 3 additional checks as above.

```{r}

```

What do these results show?

> 


Let's further investigate. In the chunk below, add a column to your stalk-eyed fly tibble with the residuals. There are a few ways to access the residuals:

- `fm$residuals`
- `resid(fm)`
- `augment()` from the `broom` package (see example in the lecture slides)

```{r}

```

Finally, plot a histogram of the *residuals* for the fitted model. Color by food source and facet the plot by food source.

```{r}

```

What do you notice about the residuals? Does this agree with the diagnostics above?

> 


## AI Statement

Please indicate whether you used any generative AI tools to help you with this problem set and explain how you used these. An example of how to structure this statement is below. If you did not use generative AI, you can simply state: "I did not use AI for this problem set."

Example statement structure: "I acknowledge the use of AI in completing this assignment and would like to provide a brief explanation of how I utilized AI, specifically [LLM], as a tool to support my work. For this assignment, I employed [LLM] to [describe the specific purpose or task]. To do so I crafted the following prompts: [List prompts used]"

>


