---
title: 'Problem Set 07'
author: "Your Name Here"
date: 'Last updated: `r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(performance)
library(see)
library(broom)
library(readxl)
library(cowplot)

theme_set(theme_classic())
```

## Setup

This problem set will use some additional packages that you might not have installed:

- "performance"
- "see"
- "ggplotr"
- "broom"
- "readxl"
- "cowplot"

If you do not have these packages installed already, go ahead and do that before you start. You can use the "Packages" pane or the function, e.g., `install.packages("performance")


## Lion noses

In PS 6, we continued analyzing the data set where lion age was related to the proportion of black coloration in the nose. The goal was to be able to predict a lion's age from only a photograph of its nose.

### Activity

Copy your code from PS 6 to import the `LionAges.csv` and make a scatterplot of age regressed on proportion of black. 

Add an OLS regression line to your plot using `geom_smooth()` following the examples in the lecture slides and PC 1.

```{r}
# FIXME

LA <- read_csv("../data/LionAges.csv", show_col_types = FALSE)

ggplot(LA, aes(y = age, x = proportion.black)) +
  geom_point(size = 4, color = "navy") +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  scale_x_continuous(limits = c(0, 1)) +
  scale_y_continuous(breaks = 1:14) +
  labs(x = "Proportion Black", y = "Age (y)")
```

Copy over the linear model you fit where age is modeled by proportion of black.

```{r}
# FIXME

fm <- lm(age ~ proportion.black,
         data = LA)
```

Pass the `lm()` object to the `check_model()` function. This will print a 5-panel figure of diagnostics. You can use the "Zoom" button in the Plots panel to make the image larger.

```{r}
# FIXME

check_model(fm)
```

We'll go through the diagnostics one-by-one.

What does the "Posterior Predictive Check" represent? Do you see any problems with it?

> The green line is the density plot of the observed y values. The blue lines are simulated values based on the model fit. This looks fine here.


What does the "Linearity" panel check? Do you see any problems with it?

> This is a check that there really is a linear relationship between the x and y variables. The line is pretty wavy, but there are also not a lot of data points. It seems flat enough.


What does the "Homogeneity of Variance" panel check? Do you see any problems with it?

> This is a check that the variation in the residuals is (relatively) constant across all the values of x. It's mostly flat here, but it does deviate away fro, horizontal at large fitted values of age.


What does the "Influential Observations" panel check? Do you see any problems with it?

> This panel shows a diagnostic for points that have an undue influence on the linear model (e.g., high "leverage"). This plot show that point #30 is influential.


Finally, run the three additional checks:

- `check_heteroskedasticity()` also spelled `check_heteroscedasticity()` 
- `check_normality()`
- `check_outliers()`

```{r}
# FIXME

check_heteroskedasticity(fm)
check_normality(fm)
check_outliers(fm)
```

What do these checks report? Do the results agree with the disagnostic plots?

> `check_heteroskedasticity()` reports signficant non-homogeniety, which appears to agree with plot #3, where the variance increases with the fitted values. `check_outliers()` reports that observation 30 is an outlier.


In class we made a plot using `geom_label()`, with the observation numbers plotted instead of points. Adapt that plot here to identify the influential observation. Include the code you wrote above to also draw the OLS regression line on the plot.

```{r}
# FIXME

LA |> mutate(ID = 1:n()) |> 
  ggplot(aes(proportion.black, age, label = ID)) +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  geom_label()
```

What does it mean to say that point #30 has a lot of influence on the regression as compared to, for example point #17?

> Point 30 is both away from the regression line and at an extreme value of proportion black. Thus it is able to exert a significant "pull" on the regression line. Point 17, falls almost exactly on the regression line. So it is not able to pull the line much at all (the observed value is right where the model predicts it to be).


Filter out observation #30 by whatever method you want. Then refit the model and print the summary.

```{r}
# FIXME

LA_2 <- LA |> slice(-which.max(LA$age))
fm_2 <- lm(age ~ proportion.black,
         data = LA_2)
summary(fm_2)
summary(fm)
```

Compare the summaries for the original model with all the data points and the new model with the outlier removed. How much do the coefficients change? Does dropping one data point change your conclusions about the ability to predict age from proportion of black in the nose?

> The slope goes from 10.6 in the first model to 8.8 in the next model, so much lower, but still really high. The intercept changes as well, but that is not as interesting. Regardless of the change in slope, proportion.black is still a signficant preditor of age. The P-value are both in the 10^-7 range.


## Stalk-eyed flies

In Problem Set 6 we continued analyzing the stalk-eyed fly data. We want to look at the diagnostics for that analysis now.

### Activity

Import the `Stalkies.csv` as in PS 6 and run the same linear model. Feel free to copy your code or the code from the key.

```{r}
# FIXME

stalk <- read_csv("../data/Stalkies.csv", show_col_types = FALSE) |> 
  mutate(food_source = factor(food_source))

fm <- lm(eye_span ~ food_source, data = stalk)
summary(fm)
```


Pass the fitted model to `check_model()` as above to plot the regression diagnostics.

```{r}
# FIXME

check_model(fm)

```

What do you notice is different about these diagnostic plots compared to the one above?

> There are only two x "values", so the plots look really wonky.


What can you conclude about the utility of diagnostic plots when the predictor is a categorical variable?

> It's not that useful.


Run the same 3 additional checks as above.

```{r}
# FIXME

check_heteroskedasticity(fm)
check_normality(fm)
check_outliers(fm)
```

What do these results show?

> These check so significant heteroscedasticity and non-normality of residuals.


Let's further investigate. In the chunk below, add a column to your stalk-eyed fly tibble with the residuals. There are a few ways to access the residuals:

- `fm$residuals`
- `resid(fm)`
- `augment()` from the `broom` package (see example in the lecture slides)

```{r}
# FIXME

stalk <- augment(fm, stalk)
```

Finally, plot a histogram of the *residuals* for the fitted model. Color by food source and facet the plot by food source.

```{r}
# FIXME

ggplot(stalk, aes(.resid, fill = food_source)) +
  geom_histogram() +
  facet_grid(food_source ~ .)
```

What do you notice about the residuals? Does this agree with the diagnostics above?

> The corn residuals are very compact and probably normally distributed. The cotton residuals are much more spread out, and it's hard to tell of they are normal or not. This seems to agree with the diagnostics: different variances and non-normal distributions of residuals.


## Working with lists

Lists are one of the most flexible data structures in R. They can hold any combination of other types of R objects (including other lists). R functions like `lm()` return lists.^[R functions can only return a single object (unlike languages like python and MATLAB, which can return multiple objects). So everything that needs to be returned from a function is placed into a list, which gets returned.]

Lists are indexed a little differently from vectors and matrices. List indices are indexed using double brackets: `[[1]]` extracts (or assigns) the first element. Run the following code:

```{r}
mylist <- list()
mylist[[3]] <- 24
mylist
```

We first create an empty list, using the `list()` function. The second line assigns 24 to the 3rd element. When we print the list (last line), you can see that the 1st and 2nd elements are created and left as `NULL`. Managing lists can be harrowing:

```{r}
mylist[[1]] <- "hello world"
mylist[[2]] <- list(x = 12, y = "y", z = seq(1, 5, by = 0.2))
mylist
```

Now the 1st element has a string, the 2nd has a list (which itself contains a numeric, a string, and a vector). How would you extract the 3rd element from the `z` vector?

```{r}
# FIXME
mylist[[2]]$z[3]
```

We will use lists in a much less convoluted way, which is what you should do until you get comfortable (once you are comfortable with lists, their immense flexibility makes them a great way to pass collections of different kinds of data around among code).

For now, we will use a list to hold 4 plots, which we will arrange in a grid to make a usable figure.


### Activity

The file `mammals.xlsx` contains data for about 50 different traits. *All* 5,416 species of mammals are included in the file, though none has data for all traits. We want to make some quick plots of four different traits plotted against body mass.

Rather than creating separate plots, we want to make them inside a loop, where we iterate through the columns of interest. This will make it easy to modify the format of the plots all at once, reducing the amount of code to keep up with. For example, change all the points to be transparent in one line rather than in four places.

1. Load the data from `mammals.xlsx`.
2. Create a column that holds the log10 transformation of `adult_body_mass_g` and one for the log10 of `basal_met_rate_m_l_o2hr`.
    - Good names might be "log_Mass" and "log_BMR"
3. Create a vector of strings that we will use to iterate through the columns for analysis (we used `cols_analyze`). It should contain the following strings:
    - `ageat_eye_opening_d`
    - `log_BMR` (or whatever you called the log10 transformation of `basal_met_rate_m_l_o2hr`)
    - `gestation_len_d`
    - `litter_size`
4. Create a vector of strings to use for the y axis labels. Make sure that the order matches the order in your vector #3.
5. Drop any rows that have missing values for log10 Body Mass.
6. Create an empty list to hold plots: `pp <- list()` for example.

This is all we need for the setup. At this point, you have the data, two vectors (one of columns names and one of y-axis labels), and an empty list.

```{r}
# FIXME
M <- read_excel("../data/mammals.xlsx", na = "NA") |>
  mutate(log_Mass = log10(adult_body_mass_g),
         log_BMR = log10(basal_met_rate_m_l_o2hr))

cols_analyze <- c("ageat_eye_opening_d",
                  "log_BMR",
                  "gestation_len_d",
                  "litter_size")

axis_labels <- c("Age at Eye Opening",
                 "log10 BMR",
                 "Gestation Length",
                 "Litter Size")

M <- M |> 
  select(log_Mass, all_of(cols_analyze)) |> 
  drop_na(log_Mass)

pp <- list()
```

No we can work on a loop to iterate through the columns, making a plot for each column. Here is the [pseudocode](https://builtin.com/data-science/pseudocode):

1. Create the framework for a `for` loop.
    - We want to iterate from 1 to `length(cols_analyze)`
2. Extract the *ii*th element of the columns to analyze vector. Assign to an object `col`.
3. Extract the *ii*th element of the y-axis label to `lab`.
4. Make a bivariate plot of the column vs. log Body Mass (assign to an object). This will look a little different from ggplots you have made before, particularly the aesthetics. Because you are supplying raw strings, you need to use the `.data` object, which is what ggplot uses internally to keep track of data. It will look something like: 
    - `aes(x = .data[["log_Mass"]], y = .data[[col]])` depending on what variable name you used for log10 Body mass
5. Assign the plot to the *ii*th element of your list.

```{r}
# FIXME
for (ii in 1:length(cols_analyze)) {
  col <- cols_analyze[ii]
  lab <- axis_labels[ii]
  
  p <- ggplot(M, aes(x = .data[["log_Mass"]], y = .data[[col]])) +
    geom_point(alpha = 0.5, size = 0.5) +
    labs(x = "log10 Body Mass (g)", y = lab) +
    theme(axis.text = element_text(size = 10),
          axis.title = element_text(size = 10))
  pp[[ii]] <- p
}
```

At the end of the loop you will have a list that contains 4 plots. Try printing a plot by running `pp[[1]]` at the command line. If you get warnings saying something like `Removed ... rows containing missing values (geom_point).`, that is just ggplot saying that there are `NA`s in the data. A nice fix-up would be to go back to the data and `drop_na()` for the *ii*th column before you plot.

Now you can make a 2x2 grid of plots. The `cowplot` function `plot_grid()` can take a list of plots as an argument: `plotlist = ...`. Pass your list of plots, and specify 2 columns and 2 rows (check the help for `?plot_grid`).

```{r}
# FIXME
plot_grid(plotlist = pp, ncol = 2, nrow = 2,
          labels = c("a", "b", "c", "d"),
          label_size = 8)
```

You will get 4 sets of warning messages about missing data. This is okay. We could edit the code to subset and filter out NAs inside the loop to avoid these messages.

Now you can fine tune the plots in the code chunk above:

- Add labels for the x and y axes. The y label is the `lab` string you extracted from your vector of y axis labels.
- Make the points smaller (`size = 0.5`) and transparent
- Change the font size on the axis labels by adding the following code to your ggplot:

```
+ theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 10))
```

- Pass the argument `labels = c("a", "b", "c", "d")` to `plot_grid()` to add labels for each plot.

As you can see rather than sequentially editing 4 sets of code, trying to make sure they all stay in sync, we can just edit one plotting statement and change the data in each iteration.
