---
title: 'Problem Set 09'
author: "Your Name Here"
date: 'Last updated: `r Sys.Date()`'
format:
  html:
    toc: true
    toc_depth: 4
    toc_float: true
    embed-resources: true
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(cowplot)
library(performance)
library(GGally)

theme_set(theme_classic())
```


## Heart Transplant Survivorship

The data in `Heart_Transplants.xlsx` contains data on survivorship (in days; `Survival_Days`) for three groups of heart transplant patients. Patients were groups by the severity of the mismatch between donor and recipient (`Mismatch_Degree`): `Low`, `Medium`, and `High`. A low *mismatch* means a *good* match.


### Activity

Load the data from the Excel file and look at the structure with either `str()` or `glimpse()`.

```{r}



```

Notice that `Mismatch_Degree` is currently a character vector. To use it in a linear model, we should convert it to a factor.^[R will do this automatically, but it is good practice to do the conversion, in this case to get the levels of the factor in the order we want.]

We could do this with R's built-in function `factor()`. The drawback is that, by default, `factor()` will order the factors in alphabetical order. For these data, the alphabetical order will put the `High` mismatch first, which is not ideal. We want to have `Low` come first.

[`forcats`](https://forcats.tidyverse.org/) is a package that greatly simplifies working with factors in R. It has functions to create factors in the order that they appear in the `data.frame`, in the order of frequency, and in any arbitrary order. `forcats` is built into the tidyverse, so you do not need to load it separately.

Because the data or organized so that the `Low` group is first, we can use `fct_inorder()` to make the factor in the sequence that we want.

Use `mutate()` and `fct_inorder()` to make `Mismatch_Degree` into a factor.

```{r}


```

There are only 39 observations total, so we may as well plot the raw data. Modify the code from the last few problem sets for plotting the points with the group means and standard errors.

```{r}


```

Describe what you see in the plots. You would like to be able to use a linear model (one-way ANOVA) to compare the mean survival days between groups. In what way does the data not appear to satisfy the assumptions of ANOVA?

> 


Find a transformation of `Survival_Days` that corrects the problems observed above. Plot your transformed data as above (copy and modify the code). Briefly explain why you chose the transformation that you chose.

```{r}


```

> 



Using `lm()`, fit a linear model in which transformed survival is modeled by the degree of mismatch. Save this to an object. Use `anova()` on that object to get the ANOVA table for the model you just fit.

```{r}



```

What do the results of this linear model show?

> 


## Bird abundance

The file `Birds.xlsx` contains data from Vuilleumier, F. (1970) Insular Biogeography in Continental Regions. I. The Northern Andes of South America. *American Naturalist* 104:373-388.

This paper explores the numbers of bird species in isolated "islands" of páramo vegetation in the Andes. The Missouri Botanical Garden can explain it best:

> Within the tropical regions of Mexico, Central and South America, Africa, Malesia including New Guinea, and Hawaii, there is a vegetation type that occurs between the upper limit of continuous, closed-canopy forest (i.e., forest line or timberline) and the upper limit of plant life (i.e., snow line) that is characterized by tussock grasses, large rosette plants, shrubs with evergreen, coriaceous and sclerophyllous leaves, and cushion plants.  This vegetation type is scattered along the crests of the highest mountain ranges or on isolated mountaintops between about 3000 meters (m) and 5000 m, like islands in a sea of forest.

We would like to use these data to see what the best predictor(s) of bird abundance is/are. The data contain species abundance (`N_Species`) and geographical information for 14 "islands". Other data include:

1. `Area`: "Island" size (thousands of square km)
1. `Elevation`: Elevation (thousands of meters)
1. `Dist_to_Ecuador`: Distance to Ecuador (km)
1. `Dist_to_Island`: Distance to nearest island (km)


### Activity

We will use multiple regression to find the relative importance of each of these variables in predicting species abundance.

Start by loading the data and plotting a scatterplot matrix (`GGally::ggscatmat()`). Think about any variables that might need transformation. There are only 14 observations, so it's going to be hard to discern normality. Here you are just looking for really obviously right or left skewed distributions.

```{r}


```

Describe any patterns you see in the raw data.

> 


What patterns do you see in the correlations in the upper triangle of the plot? Note that 1's run along the diagonal (a variable correlated to itself is 1), and that the upper and lower triangles are symmetrical. Also note that you haven't done any tests of significance, so you can't say anything definitive about "significant correlations".

> 


Fit a linear model where species abundance is predicted by the other four variables. Save this model to an R object.

```{r}


```

Use `summary()` to get information about the model fit.

```{r}


```

R^2^ is a measure of how much of the variation in the outcome variable (`N_Species`) is explained by the predictors. R^2^ for linear models in R is "Multiple R-squared" (ignore "Adjusted R-squared") in the summary.

What is the R^2^ for the regression model?

> 


Overall, are the four variables able to predict species abundance in combination (overall ANOVA)?

> 



What variable or variables are significantly different from 0? Does species abundance increase or decrease with a one unit increase in this/these variable(s)?

> 


Run `check_model()` to see if there are any concerns with the data.

```{r}


```

Does anything seem concerning?

> 
