---
title: 'Problem Set 10'
author: "Your Name Here"
date: 'Last updated: `r Sys.Date()`'
format:
  html:
    toc: true
    toc_depth: 4
    toc_float: true
    embed-resources: true
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false



ggplot2::theme_set(ggplot2::theme_classic())
```


## Predicting height from limb lengths

We should be able to predict someone's height from the lengths of their limbs. Based on measurements from a very large number of people, we know that limb length is about 40-50% of total height. So we should be able to develop a linear model for prediction.

The file `Limb_Lengths.csv` contains data for 100 heights and associated left and right limb lengths.


### Activity

Load the data for limb lengths. Try to use the function `summarize_all()` to calculate the mean and standard deviation for each column. Look at the help to figure out how to apply two different functions to each column

```{r}


```

Make two plots, one for left limb length vs. height and one for right limb length vs. height. Add an OLS regression line to each using `geom_smooth(). Use `cowplot::plot_grid()` to put them side by side.

```{r}


```

Based on the plots, what do you think the relationship between limb length and height is (positive, negative, or no relationship)?

> 


Fit a single linear model in which height is predicted by both left and right limb lengths (this is an "additive" model). Print out the summary of the model.

```{r}


```

Explain the results that you find. Why do you think that neither left nor right limb length is a good predictor of height in the model you fit? Use `ggscatmat()` and `check_model()` to help diagnose the problem.

```{r}


```

> 


Fit separate linear models in which height is predicted by left and then right limb length. Look at the summaries of these linear models.

```{r}


```

Do these models make more sense?

> 



## Neandertal Brains

There are well-known differences in both body size between modern humans and Neanderthals. But are there differences in brain size as well, adjusted for body size? Ruff and colleagues (1997) tried to answer just that question.^[Ruff, C.B., E. Trinkaus, and T.W. Holliday. 1997. Body mass and encephalization in Pleistocene *Homo*. *Nature* 387: 173-176.]

### Activity

The file `NeandertalBrainSize.csv` contains data on estimated log body mass, log brain size, and `Species`. Load the file, and then:

1. Convert `Species` to a factor.
2. Re-level `Species` so that "Recent" is the base level of the factor.
    - Or do both steps at the same time.

```{r}


```

Make a scatterplot of brain size vs. body mass, with points colored by species. 

- See if you can find information on the internet about how to move the legend from the side of the plot into the lower right hand corner.
- Add a linear regression line to the plots

```{r}


```

What is your interpretation of the plot? What patterns do you observe?

> 


Fit three linear models and assign each to an object:

1. Brain size modeled by body mass
2. Brain size modeled by body mass and species (additive)
3. Brain size modeled by body mass and species with the mass X species interaction

Then print the summary for each model.

```{r}


```

In words, describe what each regression model represents. Feel free to follow the example in the slides for units 9 and 10.

> 


Which model do the lines in the figure above represent?

> 


What does each line in the summary of Model 3 represent?

> 

