---
title: 'Problem Set 11'
author: "Your Name Here"
date: 'Last updated: `r Sys.Date()`'
format:
  html:
    toc: true
    toc_depth: 4
    toc_float: true
    embed-resources: true
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
library(tidyverse)


theme_set(theme_classic())
```


Packages you will need for this problem set:

- `AICcmodavg`
- `cvTools`

Install them now using the "Packages" pane (if you haven't already).


## Neandertal Brains

In Problem Set 10, we looked at a dataset of Neandertal brain size. Review that question.

Using the code you wrote (or the code from the key), load the `NeandertalBrainSize.csv` dataset and relevel so that "Recent" is the base level for the `Species` variable.

```{r}

```

Copy the code you wrote for making a scatterplot of the data to remind yourself of the patterns.

```{r}

```

We fit three linear models and assigned each to an object:

1. Brain size modeled by body mass
2. Brain size modeled by body mass and species (additive)
3. Brain size modeled by body mass and species with the mass X species interaction

Copy your model code from PS 10 and fit the models:

```{r}

```

We had left off at this point: fitting the models but not deciding which was the best fit to the data.


### Information criteria

Use the `aictab()` function from the `AICcmodavg` to generate the AICc model comparison table for the three models. Follow the template in the lecture slides. For example, `aictab()` expects to recieve a `list()` of model objects (with optional names).

```{r}

```

Notice that the values of AICc are negative. This can happen and is not a problem at all or a sign that something has gone wrong.

You have to think a little harder about what "lower" means (in this case *more negative*). The interpretation stays the same -- lower AICc is better.

Which model has the lowest AICc value?

>


What is the difference (Delta-AICc) between the best and 2nd best model?

>


Overall, what is your interpretation of the `Delta_AICc` column?

>


Interpret the model weights in the `AICcWt` column. Remember that model weights can be interpreted as a probability that the model in that row is the best model for out-of-sample prediction.

>


If these were your data, how would you proceed in the "Results" and "Discussion" sections of your manuscript?

>


### Cross-validation

Follow the example in the lecture slides to carry out *leave-one-out* cross validation on these models using the `cvFit()` function from the `cvTools()` package. Cross-validation uses the data to determine the root-mean-squared prediction error for a model (RMSE).

```{r}

```

What are the RMSE values for the three models?

>


What is your interpretation of the LOO cross-validation model comparison?

>


Considering both the AICc table and the cross-validation, do the results agree or disagree? Why or why not?

>


What do you interpret about the possible differences in brain scaling vs. body size in Recent humans vs. Neandertals based on these analyses?

>


## AI Statement

Please indicate whether you used any generative AI tools to help you with this problem set and explain how you used these. An example of how to structure this statement is below. If you did not use generative AI, you can simply state: "I did not use AI for this problem set."

Example statement structure: "I acknowledge the use of AI in completing this assignment and would like to provide a brief explanation of how I utilized AI, specifically [LLM], as a tool to support my work. For this assignment, I employed [LLM] to [describe the specific purpose or task]. To do so I crafted the following prompts: [List prompts used]"

>
