---
title: 'Problem Set 10'
author: "Your Name Here"
date: 'Last updated: `r Sys.Date()`'
format:
  html:
    toc: true
    toc_depth: 4
    toc_float: true
    embed-resources: true
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(readxl)
library(cowplot)
library(performance)
library(ggsci)
library(GGally)

theme_set(theme_classic())
```


## Effects of calcium and vitamin D on bone mineral density

```{r}
#| echo: false
# Simulate data
# FIXME

library(tidyverse)
library(glue)

set.seed(2389472)
n <- 500

d <- tibble(
  age = round(runif(n, 25, 75)),
  calcium = pmax(round(rnorm(n, 900, 400)), 100),
  vitamin_d = pmax(round(rnorm(n, 60, 25), 1), 10)
) |>
  mutate(
    bmd_raw = 0.8 +
      0.00005 * calcium +
      0.0008 * vitamin_d +
      0.00001 * calcium * vitamin_d -
      0.006 * age +
      rnorm(n, 0, 0.1),
    bmd = as.numeric(scale(bmd_raw))
  )

write_csv(d, file = "../data/Human_BMD.csv")
```

Bone mineral density (BMD) is a measure of the amount of mineral (primarily a type of calcium phosphate) in a small of bone. BMD measured at the hip is a primary indicator of osteoporosis and risk of fracture in a fall.

BMD follows a predictable pattern: it increases through adolescence, reaches a peak in the late twenties, remains relatively stable through middle adulthood, then declines gradually, accelerating following menopause as estrogen levels drop in those who menstruate.

Two dietary factors strongly influence this BMD trajectory: calcium provides the mineral for making bone, and vitamin D regulates calcium absorption.

In this problem, you will use simulated data loosely based on reference ranges from NHANES to model BMD as a function of age, calcium intake, vitamin D status, and their interactions.

### Activity

Import the data `Human_BMD.csv` into an R object `BMD` (This will make code later work more easily.).

You will find 5 columns:

- `age`: Age in years
- `calcium`: Dietary calcium level in mg/dL
- `vitamin_d`: Blood serum Vitamin D in nmol/L
- `bmd_raw`: Raw BMD measured by [DXA](https://my.clevelandclinic.org/health/diagnostics/10683-dexa-dxa-scan-bone-density-test) in g/cm^2^
- `bmd`: Standardized BMD (no units). The raw BMD score is standardized into a Z-score (subtracting the mean and dividing by the standard deviation) for comparison. Because these data are simulated, the BMD values here are not directly comparable to a fully calibrated clinical DXA scanner (but they are close).

```{r}
# FIXME

BMD <- read_csv("../data/Human_BMD.csv", show_col_types = FALSE)
```

Make a scatterplot of BMD vs. Age.

- Add a regression line with `geom_smooth()`
- Add axis labels.

```{r}
# FIXME

ggplot(BMD, aes(x = age, y = bmd)) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE) +
  labs(x = "Age (years)", y = "BMD (Z)")
```

What pattern do you observe in the scatterplot and the regression line? Does that pattern agree with the biological description above?

> There is an overall negative relationship between BMD and age. As people age their BMD decreases.

- Fit a linear model where BMD is predicted by age.
- Print the summary.

```{r}
# FIXME

fm <- lm(bmd ~ age, data = BMD)
summary(fm)
```

Interpret the results of the linear model. Do they support your interpretation of the plot? Do they agree with predictions based on the biology?

> There is a significant negative slope of age (BMD decreases by -0.014 for each year). The P-value is ~0.000001. This is a small effect but significant.

The code chunk below makes a plot of BMD as a function of calcium with the points colored by vitamin D level.

- You might have to install the `colorspace` package.
- Change `eval` to `true`
- Run the chunk.

```{r}
#| eval: true
# FIXME ^^^

library(colorspace)

BMD |>
  ggplot(aes(x = calcium, y = bmd, color = vitamin_d)) +
  geom_point(size = 3) +
  scale_color_continuous_diverging(
    palette = "Purple-Green",
    name = "Serum Vitamin D"
  ) +
  labs(x = "Dietary Calcium (mg/dL)", y = "BMD (Z)")
```

What pattern do you observe in the relationship between BMD and calcium? What pattern do you observe between BMD and vitamin D?

> There is a positive relationship between BMD and calcium. At any level of calcium, higher values of vitamin D are associated with higher BMD.

- Fit a linear model where BMD is predicted by calcium, vitamin D, and the calcium by vitamin D interaction. Assign this model to the object `fm2` (we will use this model below).
- Print the model summary.

```{r}
# FIXME

fm2 <- lm(bmd ~ calcium * vitamin_d, data = BMD)
summary(fm2)
```

Based on the linear model, what is your interpretation of the relationship between the two predictors and their interaction with BMD?

> All of the coefficients are small in the summary. Neither the coefficient for calcium nor vitamin D is different from zero (P = 0.809 and 0.519). The only significant coefficient is the interaction.

It's pretty difficult to figure out what the numbers mean in the regression table. We can make a plot using specific values of vitamin D to see the relationship.

- In the chunk below, change `eval` to `true`. Run the chunk to make the plot.
- Work through the code line by line and try to write comments of what you think each section is doing. Print `vd_cuts` and `grid` to see what they contain.

```{r}
#| eval: true
# FIXME ^^^^

# Find the Q1, median, and Q3 quantiles for vitamin D
vd_cuts <- quantile(BMD$vitamin_d, probs = c(0.25, 0.50, 0.75))

# Create a grid of values where calcium has 100 values between the minimum
# and maximum value.
grid <- crossing(
  calcium = seq(min(BMD$calcium), max(BMD$calcium), length.out = 100),
  vitamin_d = vd_cuts,
  age = median(BMD$age)
)

# Calculate predicted values of BMD for each row of grid.
# Make Vitamin D into a factor
grid <- grid |>
  mutate(
    predicted_bmd = predict(fm2, newdata = grid),
    vitamin_d_level = factor(
      vitamin_d,
      levels = vd_cuts,
      labels = glue(
        "{c('25th Percentile', '50th Percentile', '75th Percentile')} ({round(vd_cuts)} nmol/L)"
      )
    )
  )

# Make the plot
ggplot() +
  geom_line(
    data = grid,
    aes(x = calcium, y = predicted_bmd, color = vitamin_d_level),
    linewidth = 1
  ) +
  scale_color_manual(values = c("#B3E0A6", "#64AC59", "#24693D")) +
  labs(
    x = "Dietary Calcium (mg/dL)",
    y = "BMD (Z)",
    color = "Serum Vitamin D (nmol/L)"
  ) +
  theme_classic()
```

Try to interpret the plot in the context of the first plot of BMD vs. calcium and the linear model. Where does the significant interaction term come from?

> As calcium level increases, increased values of Vitamin D result in larger values of BMD. When calcium is low the effect is really small. This is where the significant interaction term comes from.


## Neandertal Brains

There are well-known differences in both body size between modern humans and Neanderthals. But are there differences in brain size as well, adjusted for body size? Ruff and colleagues (1997) tried to answer just that question.^[Ruff, C.B., E. Trinkaus, and T.W. Holliday. 1997. Body mass and encephalization in Pleistocene *Homo*. *Nature* 387: 173-176.]

### Activity

The file `NeandertalBrainSize.csv` contains data on estimated log body mass, log brain size, and `Species`. Load the file, and then:

- Convert `Species` to a factor.
- Re-level `Species` so that "Recent" is the base level of the factor.
    - Or do both steps at the same time.

```{r}
# FIXME
NBS <- read_csv(
  "https://raw.githubusercontent.com/QM-Life-Sci/QMLS_1/refs/heads/main/data/NeandertalBrainSize.csv",
  col_types = "ddc"
) |>
  mutate(Species = fct_relevel(Species, "Recent"))
```

Make a scatterplot of brain size vs. body mass, with points colored by species.

- See if you can find information on the internet about how to move the legend from the side of the plot into the lower right hand corner.
- Add a linear regression line to the plots

```{r}
# FIXME

NBS |>
  ggplot(aes(ln_Mass, ln_Brain, color = Species)) +
  geom_point(size = 3) +
  geom_smooth(formula = y ~ x, se = FALSE, method = "lm") +
  scale_color_d3() +
  guides(color = guide_legend(position = "inside")) +
  theme(
    legend.position = "inside",
    legend.justification.inside = c(1, 0.05)
  ) +
  labs(x = "ln Body Mass", y = "ln Brain Size")

# Alternately
# NBS |> ggplot(aes(ln_Mass, ln_Brain, color = Species)) +
#   geom_point(size = 3) +
#   geom_smooth(formula = y ~ x, se = FALSE, method = "lm") +
#   scale_color_d3() +
#   guides(color = guide_legend(position = "inside")) +
#   theme(legend.position = "inside",
#         legend.position.inside = c(0.75, 0.12)) +
#   labs(x = "ln Body Mass", y = "ln Brain Size")
```

What is your interpretation of the plot? What patterns do you observe?

> (log) brain size increases with (log) body mass in both Recent human and Neandertals. For a given body size, Recent humans appear to have larger brains.


Fit three linear models and assign each to an object:

- Brain size modeled by body mass
- Brain size modeled by body mass and species (additive)
- Brain size modeled by body mass and species with the mass X species interaction

Then print the summary for each model.

```{r}
# FIXME

fm1 <- lm(ln_Brain ~ ln_Mass, data = NBS)
fm2 <- lm(ln_Brain ~ ln_Mass + Species, data = NBS)
fm3 <- lm(ln_Brain ~ ln_Mass * Species, data = NBS)

summary(fm1)
summary(fm2)
summary(fm3)
```

In words, describe what each regression model represents. Feel free to follow the example in the slides for units 9 and 10.

> Model 1: A single regression line (Brain size vs. Mass) passes through all points, regardless of the Species. Model 2: Separate intercepts are fit for both Species, but a single slope is shared between the two. The lines are constrained to be parallel. Model 3: Separate slopes and intercepts for both Species. Essentially this fits two separate lines.

Which model do the lines in the figure above represent?

> Model 3 with the interaction.

What does each line in the summary of Model 3 represent?

> (intercept): The intercept for the Recent group. ln_Mass: The slope for the recent group. SpeciesNeanderthal: The change in (Intercept) for being in the Neandertal group. ln_Mass:SpeciesNeanderthal: The change in the ln_Mass slope for being in the Neandertal group.


## AI Statement

Please indicate whether you used any generative AI tools to help you with this problem set and explain how you used these. An example of how to structure this statement is below. If you did not use generative AI, you can simply state: "I did not use AI for this problem set."

Example statement structure: "I acknowledge the use of AI in completing this assignment and would like to provide a brief explanation of how I utilized AI, specifically [LLM], as a tool to support my work. For this assignment, I employed [LLM] to [describe the specific purpose or task]. To do so I crafted the following prompts: [List prompts used]"

>
