---
title: 'Problem Set 10'
author: "Your Name Here"
date: 'Last updated: `r Sys.Date()`'
format:
  html:
    toc: true
    toc_depth: 4
    toc_float: true
    embed-resources: true
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

```


## Effects of calcium and vitamin D on bone mineral density

Bone mineral density (BMD) is a measure of the amount of mineral (primarily a type of calcium phosphate) in a small of bone. BMD measured at the hip is a primary indicator of osteoporosis and risk of fracture in a fall.

BMD follows a predictable pattern: it increases through adolescence, reaches a peak in the late twenties, remains relatively stable through middle adulthood, then declines gradually, accelerating following menopause as estrogen levels drop in those who menstruate.

Two dietary factors strongly influence this BMD trajectory: calcium provides the mineral for making bone, and vitamin D regulates calcium absorption.

In this problem, you will use simulated data loosely based on reference ranges from NHANES to model BMD as a function of age, calcium intake, vitamin D status, and their interactions.


### Activity

Import the data `Human_BMD.csv` into an R object `BMD` (This will make code later work more easily.).

You will find 5 columns:

- `age`: Age in years
- `calcium`: Dietary calcium level in mg/dL
- `vitamin_d`: Blood serum Vitamin D in nmol/L
- `bmd_raw`: Raw BMD measured by [DXA](https://my.clevelandclinic.org/health/diagnostics/10683-dexa-dxa-scan-bone-density-test) in g/cm^2^
- `bmd`: Standardized BMD (no units). The raw BMD score is standardized into a Z-score (subtracting the mean and dividing by the standard deviation) for comparison. Because these data are simulated, the BMD values here are not directly comparable to a fully calibrated clinical DXA scanner (but they are close).

```{r}

```

Make a scatterplot of BMD vs. Age.

- Add a regression line with `geom_smooth()`
- Add axis labels.

```{r}

```

What pattern do you observe in the scatterplot and the regression line? Does that pattern agree with the biological description above?

>

- Fit a linear model where BMD is predicted by age.
- Print the summary.

```{r}

```

Interpret the results of the linear model. Do they support your interpretation of the plot? Do they agree with predictions based on the biology?

>

The code chunk below makes a plot of BMD as a function of calcium with the points colored by vitamin D level.

- You might have to install the `colorspace` package.
- Change `eval` to `true`
- Run the chunk.

```{r}
#| eval: false

library(colorspace)

BMD |>
  ggplot(aes(x = calcium, y = bmd, color = vitamin_d)) +
  geom_point(size = 3) +
  scale_color_continuous_diverging(
    palette = "Purple-Green",
    name = "Serum Vitamin D"
  ) +
  labs(x = "Dietary Calcium (mg/dL)", y = "BMD (Z)")
```

What pattern do you observe in the relationship between BMD and calcium? What pattern do you observe between BMD and vitamin D?

> 

- Fit a linear model where BMD is predicted by calcium, vitamin D, and the calcium by vitamin D interaction. Assign this model to the object `fm2` (we will use this model below).
- Print the model summary.

```{r}

```

Based on the linear model, what is your interpretation of the relationship between the two predictors and their interaction with BMD?

>

It's pretty difficult to figure out what the numbers mean in the regression table. We can make a plot using specific values of vitamin D to see the relationship.

- In the chunk below, change `eval` to `true`. Run the chunk to make the plot.
- Work through the code line by line and try to write comments of what you think each section is doing. Print vd_cuts and grid to see what they contain.

```{r}
#| eval: false

#
vd_cuts <- quantile(BMD$vitamin_d, probs = c(0.25, 0.50, 0.75))

#
grid <- crossing(
  calcium = seq(min(BMD$calcium), max(BMD$calcium), length.out = 100),
  vitamin_d = vd_cuts,
  age = median(BMD$age)
)

#
grid <- grid |>
  mutate(
    predicted_bmd = predict(fm2, newdata = grid),
    vitamin_d_level = factor(
      vitamin_d,
      levels = vd_cuts,
      labels = glue(
        "{c('25th Percentile', '50th Percentile', '75th Percentile')} ({round(vd_cuts)} nmol/L)"
      )
    )
  )

#
ggplot() +
  geom_line(
    data = grid,
    aes(x = calcium, y = predicted_bmd, color = vitamin_d_level),
    linewidth = 1
  ) +
  scale_color_manual(values = c("#B3E0A6", "#64AC59", "#24693D")) +
  labs(
    x = "Dietary Calcium (mg/dL)",
    y = "BMD (Z)",
    color = "Serum Vitamin D (nmol/L)"
  ) +
  theme_classic()
```

Try to interpret the plot in the context of the first plot of BMD vs. calcium and the linear model. Where does the significant interaction term come from?

>


## Neandertal Brains

There are well-known differences in both body size between modern humans and Neanderthals. But are there differences in brain size as well, adjusted for body size? Ruff and colleagues (1997) tried to answer just that question.^[Ruff, C.B., E. Trinkaus, and T.W. Holliday. 1997. Body mass and encephalization in Pleistocene *Homo*. *Nature* 387: 173-176.]

### Activity

The file `NeandertalBrainSize.csv` contains data on estimated log body mass, log brain size, and `Species`. Load the file, and then:

1. Convert `Species` to a factor.
2. Re-level `Species` so that "Recent" is the base level of the factor.
    - Or do both steps at the same time.

```{r}

```

Make a scatterplot of brain size vs. body mass, with points colored by species.

- See if you can find information on the internet about how to move the legend from the side of the plot into the lower right hand corner.
- Add a linear regression line to the plots

```{r}

```

What is your interpretation of the plot? What patterns do you observe?

>


Fit three linear models and assign each to an object:

1. Brain size modeled by body mass
2. Brain size modeled by body mass and species (additive)
3. Brain size modeled by body mass and species with the mass X species interaction

Then print the summary for each model.

```{r}

```

In words, describe what each regression model represents. Feel free to follow the example in the slides for units 9 and 10.

>


Which model do the lines in the figure above represent?

>


What does each line in the summary of Model 3 represent?

>
